Species persistence under climate and fishing

### 2014-01-30 Writing/MS.tex
rearranged order of problems / results so that synergy comes before critical rates

There are many stressors that can disturb an ecosystem. Ecologists have quantified the effects of a number of stressors individually \citep{Wilcoveetal1998, Crainetal2008, DarlingCote2008}. However, disturbances rarely occur in isolation, so it is important to understand how a system is affected by multiple disturbances together \citep{DoakMorris2010, Fordhametal2013, Foltetal1999}. A perturbation that has little effect when it occurs individually may still amplify the disturbance caused by a coincident perturbation \citep{Crainetal2008, DarlingCote2008}. Synergistic interactions between stressors can be especially problematic if an animal population can persist in the face of a single stressor but the combination of multiple stressors suffices to drive the population extinct (i.e. \citet{Pelletieretal2006}). However it remains difficult to predict when disturbances are likely to interact in ways not anticipated when perturbations are measured individually.

Climate change is a prime example of multiple, possibly interacting, disturbances. Species that are likely to undergo (or already undergoing) shifts in range and phenology are subject to many other disturbances including pollution, habitat fragmentation, over-exploitation, and invasive species \citep{Wilcoveetal1998, Salaetal2000, MEA2005}. A common approach to predicting future population distrubtions under climate change has been to use bioclimatic-envelope models (also known as species distribution models -- SDMs). These statistical models typically correlate presence-absence data with biophysical characteristics such as mean or maximum temperatures, rainfall, or salinity, to explain and predict how species ranges' will differ under climate change \citep{Elithetal2006, GuisanThuiller2005, GuisanZimmerman2000}. Despite these models' widespread adoption, SDMs have frequently been criticized as oversimplified as they lack species interactions, dispersal and reproductive processes \citep{KearneyPorter2009, Zarnetskeetal2012, Robinsonetal2011}. Recent work on range shifts has addressed some of these gaps by explicitly including dispersal and reproduction \citep{Berestyckietal2009, ZhouKot2011}. However these models only address one disturbance: that of climate-driven range shifts. Not only does climate change disturb ecological systems via range shifts (for example), but these range shifts may interact with anthropogenetic stressors already present. Marine fish, for example, are already moving \citep{Perryetal2005, HiddinkHoftstede2008, Rijnsdorpetal2009, Dulvyetal2008, Simpsonetal2011} and projected to move more  \citep{Kelletal2005, Mackenzieetal2007}. These organisms not only are faced with ocean acidification and temperature driven range shifts \citep{Pinskyetal2013, Barryetal1995, Nyeetal2009}, but are often harvested commercially.

Climate change and fishing have been identified as the two largest human impacts on the ocean \citep{Halpernetal2008} so there is a need to understand whether and how these disturbances will interact in the future. Predictions for how climate change will drive shifts in marine systems have focused on how physiological limitations \citep{PortnerKnust2007} and global fish distributions \citep{Cheungetal2008, Cheungetal2009, Hareetal2010, Lenoiretal2011} may change. Work considering the joint impacts of climate and fishing often consider climate fluctuations (large anomalies around the mean) rather than directional changes in climate \citep{WaltersParma1996, KingMcFarlane2006}. Work that does consider the effects of climate-driven range shifts on fishing typically are case-specific and detailed, integrating multiple drivers and disturbances \citep{Cheungetal2010, Lindegrenetal2010, Brownetal2010, Merinoetal2010, Merinoetal2010b, Plaganyietal2011, Ainsworthetal2011, Zhangetal2011, Barangeetal2011, Howardetal2013}. These predicted impacts are important for management and conservation planning \citep{Allisonetal2009}, however these models are so complex that understanding the relative importance of particular drivers, disturbances, and interactions is difficult (but see \citet{Nyeetal2013} for an approach using ecosystem-level models to discern relative importance of disturbances). Here we develop models of intermediate complexity \citep[sensu][]{Gaylordetal2005} that can help clarify the importance of two interacting disturbances: temperature-driven range shifts and harvest. Empirical case studies have identified interactions between overfishing and temperature-driven range shifts that suggest synergy in the magnitude of the disturbance \citep{Lingetal2009} and have demonstrated synergistic effects in micosm experiments between warming temperatures, harvesting and connectivity \citep{Moraetal2007}. This empirical work underscores the importance of understanding how range shifts and harvesting interact and, to our knowledge, there has been limited work examining interactions between harvesting and range shifts.

Here we construct an analytical model of a fish population subject to both harvesting and climate change-induced range-shift. We explicitly include two mechanistic processes central to species' responses to climate and fishing: reproduction and dispersal. Previous work has highlighted the importance of these two processes and their vulnerability to climate change \citep{Fordhametal2013, Hastingsetal2005}. We focus on determining how climate-driven range shifts and fishing interact to affect population persistence. We find that these disturbances interact additively, not multiplicatively which is often the default assumption when considering mortality.

@@

\noindent where $h$ is the proportion of adults harvested, $f(n)$ is the recruitment function giving the number of 
offspring produced by a population of size $n$ (accounting for density dependence), $k(x-y)$ is the dispersal kernel giving the probability of a 
larva traveling from position $y$ to position $x$, $L$ is the length of the patch, and $c$ is the rate at which it 
shifts across space. We provide a list of variables and functions in Table 1.  As we explain below, the functional form of the 
recruitment function only affects persistence through how quickly recruitment increases when the population 
size is near (but above) $0$, which is equivalent to the intrinsic growth rate, $R_0=f'(0)$. We therefore do not specify a 
recruitment function here. Similarly, the population's ability to persist only depends on how quickly harvesting 
increases when a small number of fish are introduced where they were previously absent, $h'(0)$. We 
therefore only considered a proportional harvesting function initially.

If a population of fish is going to persist in the face of a shifting environment, it must move along with the patch 
 in which it can reproduce. At equilibrium, the population will be described by a traveling wave, 
where the density of fish at a given point in space will change but the density of fish at a location relative to the 
shifting  patch will not. We therefore seek to describe how the population is distributed over the viable 
patch as it shifts through the world. We write this as $n^*(\bar{x})$, the density of fish at each point $\bar{x}$ in 
the shifting frame $\left[-\frac{L}{2}, \frac{L}{2}\right]$ As in \citet{ZhouKot2011}, the traveling wave $n^*$ must 
satisfy

@@

One solution to Equation \ref{traveling_pulse} is the `trivial' traveling pulse, $n^*(\bar{x}) = 0$ for all $x \in \left[-\frac{L}
{2}, \frac{L}{2}\right]$, i.e.~a patch with no fish in it. If a population becomes very small (or if we introduce a small 
population), one of two things can happen. First, the population may crash and the trivial traveling pulse 
without any fish may appear again. Second, those small numbers may increase and form a stable population. 
In this sense, a small population can be thought of as a perturbation to the trivial traveling pulse. If the trivial 
pulse is stable, the system will return to the trivial pulse even after a perturbation in the form of the introduction 
of a small population. If a population is to persist, even when it is small it must be able to avoid extinction and 
grow. For this to be the case, the trivial pulse must be unstable to small perturbations.

We would like to know the rate of environmental shift and the harvesting rate such that as long as the 
environment moves more slowly or we harvest less severely than those parameters, then the population will be able to persist. We call these, respectively, the critical rate of environmental shift, $c^*$, and the critical harvesting rate, $h^*$. 
We found these rates by finding the parameters that make the trivial pulse unstable. Evaluating stability in this kind of model is in  general difficult to do analytically. It becomes easier if the dispersal kernel is separable into its dependence on the 
source of larvae and its dependence on the destination of the larvae,
i.e.~if there are functions $a_i, b_i$ such that $k(x- y) = \sum^\infty_{i=1} a_i(x)b_i(y)$.  In our analyses, we used one such separable kernel, the Gaussian kernel given by

For each kernel, the population's ability to persist depends on properties of the population itself-- the expected distance a larva disperses ($\langle d \rangle$) and the intrinsic growth rate ($R_0$); properties of the environment-- the length of the viable patch ($L$) and how quickly the environment is shifting ($c$); and the harvesting rate ($h
$). If the environment shifts more quickly than the critical rate $c^*$ or the population is harvested at more than 
the critical rate $h^*$ then the population will not be able to persist, as described in the Appendix.  For a Gaussian kernel, the critical rates $c^*$ and $h^*$ are those values of $c$ and $h$ such that 
$$R_0(1-h)2\sqrt{2}\exp\left(\frac{-c^2}{8D}\right)\left[\text{erf}\left(\frac{L-c}{2\sqrt{2D}}\right)-\text{erf}\left(\frac{-L-c}{2\sqrt{2D}}\right)\right]=1.$$
A similar expression for a sinusoidal kernel is derived in the appendix.  For both kernels, the critical harvesting proportion can be approximated by a function that looks like 
\begin{equation}
h^*\sim1- \frac{1}{R_0}\cdot C(L,R_0)f(\sigma^2,c^2,L^2+3c^2)
\end{equation}
where $C(L,R_0)$ is a decreasing function of the length of the viable patch and the intrinsic growth rate.

@@

\noindent Equation \ref{recursive_m} allowed us to find the values of $m_i$ numerically. We then found the total biomass in the 
equilibrium traveling pulse by using these $m_i$ and integrating Equation \ref{separable_integrodifference}.  Whereas population persistence does not depend on the functional form of recruitment, $f$, equilibrium biomass does depend on what recruitment function we use.  We chose to use a Beverton-Holt recruitment function,

@@

\noindent Second, we examined harvesting rules more complex than harvesting a constant proportion of the population. Whereas population persistence in the analytical model does not depend on the functional form of recruitment $f$, to perform simulations we must 
specify a recruitment function.  Again, we chose to use a Beverton-Holt function.  In the first generation, we seeded the world with 50 individuals at a single point, as in \citep{ZhouKot2011}. We first ran through 150 generations in order for the population to reach equilibrium without harvesting or climate shift.  We then added harvesting pressure, allowed the population to again reach equilibrium, and finally added climate change by moving the viable patch.

We added harvesting pressure by harvesting a constant proportion of the population, in order to confirm our analytical results. We extended our work by evaluating the effect of a threshold harvest rule and marine protected 
areas (MPAs). With a threshold rule, we evaluated the population at each point in space to determine how 
much harvesting should occur. If the population abundance was below the designated threshold, no 
harvesting occurred. If the population exceeded the threshold, then a proportion of the `surplus' individuals 
were harvested

Equilibrium is calculated as the mean of 300 time steps once the difference in biomass between time step $t$ and $t+1$ was no greater than $0.1$. Equilibrium biomass and catch are defined as the total amount of fish present and harvested, respectively, at equilibrium.

@@

We wanted to understand whether the two stressors interact 
additively, synergistically, or antagonistically \citep{Crainetal2008} for both our analytical model and simulations. In order to quantify the effect of the 
stressors, we found the total biomass of the population when it reached an equilibrium traveling pulse and 
compared this equilibrium biomass in the presence and absence of each stressor individually or the two stressors together.

@@

@@

We extended a spatial to study a fish population subject to both the climate-driven range shifts and harvesting. Previous work has highlighted the importance of reproduction and dispersal to the vulnerability of fish populations to climate change \citep[\citet{Fordhametal2013}]{Hastingsetal2005} and our model explicitly includes these processes.  In particular, we derived expressions for critical harvesting rates and critical rates of environmental shift for a separable Gaussian kernel and found these critical rates using numerical simulations of a Laplace kernel.  For each kernel we studied, we found that the higher the growth rate and the better the mean dispersal distance matches the rate of environmental shift, the better a population can adjust to harvest and climate change.  More interestingly, we found a negative relationship between the critical harvesting rate and the rate of environmental shift.  That is, the more quickly the environment shifts the less harvesting it takes to drive the population extinct.  The curved line separating parameters that will allow the population to persist from those that won't is on indication of an interaction between the stressors.

>>>>

There are many stressors that can disturb an ecosystem. Ecologists have quantified the effects of a number of stressors individually \citep{Wilcoveetal1998, Crainetal2008, DarlingCote2008}.  If disturbances interact synergistically, a perturbation that has little effect when it occurs individually may amplify the disturbance caused by a coincident perturbation \citep{Crainetal2008, DarlingCote2008,Nyeetal2013,Gurevitchetal2000}.   In the most extreme (and worrying) cases, synergistic interactions will lead multiple stressors to drive a population extinct when it could persist in the face of any single stressor (i.e. \citet{Pelletieretal2006}).  If disturbances interact antagonistically, on the other hand, the effects of multiple stressors may be less than that predicted by any stressor individually.  Since disturbances rarely occur in isolation, it is important to measure the synergy between  disturbances in order to understand how a system will be affected by their presence and when multiple disturbances will drive a population extinct  \citep{DoakMorris2010, Fordhametal2013, Foltetal1999}.

Climate change and fishing have been identified as the two largest human impacts on the ocean \citep{Halpernetal2008}.  They therefore present an important case study of how disturbances interact in their effects on biological populations  and understanding these interactions will be crucial to managing populations subjected to these disturbances.  In response to climate change, marine fish are already moving \citep{Perryetal2005, HiddinkHoftstede2008, Rijnsdorpetal2009, Dulvyetal2008, Simpsonetal2011} and projected to move more in the future \citep{Kelletal2005, Mackenzieetal2007}.  Species that are likely to undergo or already undergoing shifts in range are also subject harvesting, in addition to many other disturbances including pollution, ocean acidification, habitat fragmentation, and invasive species \citep{Wilcoveetal1998, Salaetal2000, MEA2005, Pinskyetal2013, Barryetal1995, Nyeetal2009}.  Empirical case studies have identified interactions between overfishing and temperature-driven range shifts that suggest synergy in the magnitude of the disturbance \citep{Lingetal2009} and have demonstrated synergistic effects in microcosm experiments between warming temperatures, harvesting and connectivity \citep{Moraetal2007}. This empirical work underscores the importance of understanding how range shifts and harvesting interact.

A common approach to predicting future population distributions under climate change has been to use bioclimatic-envelope models (also known as species distribution models -- SDMs). These statistical models typically correlate presence-absence data with biophysical characteristics such as mean or maximum temperatures, rainfall, or salinity, to explain and predict how species ranges' will differ under climate change \citep{Elithetal2006, GuisanThuiller2005, GuisanZimmerman2000}. Despite these models' widespread adoption, SDMs have frequently been criticized as oversimplified as they lack species interactions, dispersal and reproductive processes \citep{KearneyPorter2009, Zarnetskeetal2012, Robinsonetal2011}.  Recent work on range shifts has addressed some of these gaps by explicitly including dispersal and reproduction \citep{Berestyckietal2009, ZhouKot2011}. However these models only address one disturbance: that of climate-driven range shifts.

Work considering the joint impacts of climate and fishing often consider climate fluctuations (large anomalies around the mean) rather than directional changes in climate \citep{WaltersParma1996, KingMcFarlane2006}. When the effects of climate-driven range shifts on fishing are considered, the models are typically case-specific and detailed, integrating multiple drivers and disturbances \citep{Cheungetal2010, Lindegrenetal2010, Brownetal2010, Merinoetal2010, Merinoetal2010b, Plaganyietal2011, Ainsworthetal2011, Zhangetal2011, Barangeetal2011, Howardetal2013}. These predicted impacts are important for management and conservation planning \citep{Allisonetal2009}, however these models are so complex that understanding the relative importance of particular drivers, disturbances, and interactions is difficult (but see \citet{Nyeetal2013} for an approach using ecosystem-level models to discern relative importance of disturbances). Here we develop models of intermediate complexity \citep[sensu][]{Gaylordetal2005} that can help clarify the importance of two interacting disturbances: temperature-driven range shifts and harvest.

Here we construct an analytical model of a fish population subject to both harvesting and climate change-induced range-shift. We explicitly include two mechanistic processes central to species' responses to climate and fishing: reproduction and dispersal. Previous work has highlighted the importance of these two processes and their vulnerability to climate change \citep{Fordhametal2013, Hastingsetal2005}.  We find that climate-driven range shifts and fishing interact synergistically, when their effects are measured by absolute drop in biomass.  We also find how the threshold level of harvesting that suffices to drive the population extinct depends on quickly the range is shifting.

@@

\noindent where $h$ is the proportion of adults harvested, $f(n)$ is the recruitment function giving the number of 
offspring produced by a population of size $n$ (accounting for density dependence), $k(x-y)$ is the dispersal kernel giving the probability of a 
larva traveling from position $y$ to position $x$, $L$ is the length of the patch, and $c$ is the rate at which it 
shifts across space.   Analyzing this kind of model becomes easier if the dispersal kernel is separable into its dependence on the source of larvae and its dependence on the destination of the larvae, i.e.~if there are functions $a_i, b_i$ such that $k(x- y) = \sum^\infty_{i=1} a_i(x)b_i(y)$ so we use such a kernel in our analytical expressions.
The critical parameter of the recruitment function is how quickly recruitment increases when the population 
size is near (but above) $0$, which is equivalent to the intrinsic growth rate, $R_0=f'(0)$.  Since the population's ability to persist only depends on how quickly harvesting increases when a small number of fish are introduced where they were previously absent, $h'(0)$. We 
therefore only considered a proportional harvesting function initially.  We provide a list of variables and functions in Table 1.

At equilibrium, the population will be described by a traveling wave, 
where the density of fish at a given point in space will change but the density of fish at a location relative to the 
shifting  patch will not. We sought to describe how the population is distributed over the viable 
patch as it shifts through the world in order to study the size of the population at equilibrium and whether or not the population could persist. We wrote this as $n^*(\bar{x})$, the density of fish at each point $\bar{x}$ in 
the shifting frame $\left[-\frac{L}{2}, \frac{L}{2}\right]$ As in \citet{ZhouKot2011}, the traveling wave $n^*$ must 
satisfy

@@

@@

\noindent Equation \ref{recursive_m} allowed us to find the values of $m_i$ numerically. We then found the total biomass in the 
equilibrium traveling pulse by using these $m_i$ and integrating Equation \ref{separable_integrodifference}.  We chose to use a Beverton-Holt recruitment function,

@@

@@

We wanted to understand whether the two stressors interact 
additively, synergistically, or antagonistically \citep{Crainetal2008}. In order to quantify the effect of the 
stressors, we found the total biomass of the population when it reached an equilibrium traveling pulse and 
compared this equilibrium biomass in the presence and absence of each stressor individually or the two stressors together.

@@

\subsection{Persistence }
One possible equilibrium traveling wave that solves Equation \ref{traveling_pulse} is the `trivial' traveling pulse, $n^*(\bar{x}) = 0$ for all $x \in \left[-\frac{L}
{2}, \frac{L}{2}\right]$, i.e.~a patch with no fish in it.  If a population becomes very small (or if we introduce a small 
population), one of two things can happen. First, the population may crash and the trivial traveling pulse 
without any fish may appear again. Second, those small numbers may increase and form a stable population. 
In this sense, a small population can be thought of as a perturbation to the trivial traveling pulse. If the trivial 
pulse is stable, the system will return to the trivial pulse even after a perturbation in the form of the introduction 
of a small population. If a population is to persist, even when it is small it must be able to avoid extinction and 
grow. For this to be the case, the trivial pulse must be unstable to small perturbations.

We would like to know the rate of environmental shift and the harvesting rate such that as long as the 
environment moves more slowly or we harvest less severely than those parameters, then the population will be able to persist. We call these, respectively, the critical rate of environmental shift, $c^*$, and the critical harvesting rate, $h^*$. 
We found these rates by finding the parameters that make the trivial pulse unstable.   In our analyses, we used a  separable kernel, the Gaussian kernel given by

For each kernel, the population's ability to persist depends on properties of the population itself-- the expected distance a larva disperses ($\langle d \rangle$) and the intrinsic growth rate ($R_0$); properties of the environment-- the length of the viable patch ($L$) and how quickly the environment is shifting ($c$); and the harvesting rate ($h
$).  Whereas equilibrium biomass depends on the function form of recruitment, population persistence only depends on the intrinsic growth rate $R_0$.  If the environment shifts more quickly than the critical rate $c^*$ or the population is harvested at more than 
the critical rate $h^*$ then the population will not be able to persist, as described in the Appendix.  For a Gaussian kernel, the critical rates $c^*$ and $h^*$ are those values of $c$ and $h$ such that 
$$R_0(1-h)2\sqrt{2}\exp\left(\frac{-c^2}{8D}\right)\left[\text{erf}\left(\frac{L-c}{2\sqrt{2D}}\right)-\text{erf}\left(\frac{-L-c}{2\sqrt{2D}}\right)\right]=1.$$
A similar expression for a sinusoidal kernel is derived in the appendix.  For both kernels, the critical harvesting proportion can be approximated by a function that looks like 
\begin{equation}
h^*\sim1- \frac{1}{R_0}\cdot C(L,R_0)f(\sigma^2,c^2,L^2+3c^2)
\end{equation}
where $C(L,R_0)$ is a decreasing function of the length of the viable patch and the intrinsic growth rate.

\noindent Second, we examined harvesting rules more complex than harvesting a constant proportion of the population. Whereas population persistence in the analytical model does not depend on the functional form of recruitment $f$, to perform simulations we must 
specify a recruitment function.  Again, we chose to use a Beverton-Holt function.  In the first generation, we seeded the world with 50 individuals at a single point, as in \citep{ZhouKot2011}. We first ran through 150 generations in order for the population to reach equilibrium without harvesting or climate shift.  We then added harvesting pressure, allowed the population to again reach equilibrium, and finally added climate change by moving the viable patch.  Equilibrium is calculated as the mean of 300 time steps once the difference in biomass between time step $t$ and $t+1$ was no greater than $0.1$.  Equilibrium biomass and catch are defined as the total amount of fish present and harvested, respectively, at equilibrium.

We added harvesting pressure by harvesting a constant proportion of the population, in order to confirm our analytical results. We extended our work by evaluating the effect of a threshold harvest rule and marine protected 
areas (MPAs). With a threshold rule, we evaluated the population at each point in space to determine how 
much harvesting should occur. If the population abundance was below the designated threshold, no 
harvesting occurred. If the population exceeded the threshold, then a proportion of the `surplus' individuals 
were harvested.

@@

We extended a spatial to study a fish population subject to both the climate-driven range shifts and harvesting. Previous work has highlighted the importance of reproduction and dispersal to the vulnerability of fish populations to climate change \citep[\citet{Fordhametal2013}]{Hastingsetal2005} and our model explicitly includes these processes.  To quantify the interaction between the stressors, we measured the synergy between their effects on population biomass.  We found positive synergy between the stressors and that the synergy is greatest in the region of parameter space where the equilibrium biomass is smallest.  We found similar results from the analytically derived biomass and the simulation derived biomass. This indicates that this result is robust to changes in the dispersal kernel.  We chose to measure the effect of each stressor by the absolute drop in biomass caused by the stressor, and we used the sum of the individual effects for our null prediction of the effect of both stressors, as in \citep{Crainetal2008, DarlingCote2008,Nyeetal2013}.  We could also have measured the effect by the percentage drop caused by the stressor(s) and used a multiplicative null prediction for the effect of both stressors.  In general, measuring synergy against an additive null prediction is more conservative than measuring synergy multiplicatively: the presence of additive synergy implies multiplicative synergy, but not vice versa \citep{Crainetal2008, Foltetal1999}.  Since we found small levels of positive additive synergy between the two stressors, other measures of synergy might show even higher levels of interaction.

We also derived expressions for critical harvesting rates and critical rates of environmental shift for a separable Gaussian kernel and found these critical rates using numerical simulations of a Laplace kernel.  For each kernel we studied, we found that the higher the growth rate and the better the mean dispersal distance matches the rate of environmental shift, the better a population can adjust to harvest and climate change.  More interestingly, we found a negative relationship between the critical harvesting rate and the rate of environmental shift.  That is, the more quickly the environment shifts the less harvesting it takes to drive the population extinct.  The curved line separating parameters that will allow the population to persist from those that won't is another indication of an interaction between the stressors.
